---
title: "first_plots"
author: "Marcelo Perez Guzman"
date: "2026-06-22"
output:
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{css, echo=FALSE}
.Rtable1 { 
  display: block;
  max-height: 600px;
  overflow-y: scroll;
  overflow-x: scroll;
}
```

## CHS Project, Table plotting

This is the initial table plots of the CHS project. 

#### Setup 
```{r start, message=FALSE}
# install tableone
library(tableone)
library(table1)

# get the dataset
getwd()
chs <- my_data <- read.csv("CHS_cohortE(in).csv")
str(chs)
table(chs$race)
chs$male <- factor(chs$male, levels=c(0,1), labels=c("Female", "Male"))
chs$hisp <- factor(chs$hisp, levels=c("Hispanic", "Non-Hispanic", "Unknown or Missing"))
chs$race <- factor(chs$race, levels=c("African American", "Asian", "Caucasian", "Mixed", "Others", "Unknown or Missing"))
```

## Table 1

This is what shows up after creating a table with the tableone package, which has the mean and the standard deviation of the variables

```{r tableone_pck, echo=FALSE}
# Tableone package, includes mean and standard deviation
myvars <- c("age", "male", "race", "hisp", "bmi", "educ", "pm25", "sulfate", "nitrate", "wheeze","fev1", "fvc")
CreateTableOne(data = chs, vars = myvars)
```

This is using the table1 package instead of tableone, creates a publication ready table

This table is on the demographics of the participants in the study
```{r table1_pck, echo=FALSE}
# table1 package, creates better visual tables

label(chs$age) <- "Age"
label(chs$male) <- "Sex"
label(chs$race) <- "Race"
label(chs$hisp) <- "Hispanicity"
label(chs$bmi) <- "Body-Mass Index"
label(chs$educ) <- "Education"
label(chs$pm25) <- "PM25"
label(chs$sulfate) <- "Sulfate"
label(chs$nitrate) <- "Nitrate"
label(chs$wheeze) <- "Wheeze"
label(chs$fev1) <- "FEV1"
label(chs$fvc) <- "FVC"
label(chs$ec) <- "Elemental Carbon"
label(chs$dust) <- "Dust"

table1(~ age + male + race + hisp + bmi + educ | townabbr, data=chs, render.continuous = c(. = "Mean (SD)"))


# + pm25 + sulfate + nitrate + wheeze + fev1 + fvc

```





And this is a table of the polution levels in each town
```{r, echo=FALSE}
table1(~ pm25 + sulfate + nitrate + wheeze + fev1 + fvc + dust + ec | townabbr, data=chs, render.continuous = c(. = "Mean (SD)"))
```

## Stratified stuff

Things to stratify by:
- race/hisp + lung capacity
- ETS secondhand smoke + lung capacity
- BMI + lung capacity
- race + BMI
- race + secondhand smoke
- plants + lung capacity
- homebuilt + race/hisp
- homebuilt + lung capacity